---
title: "Reproducible Research: Peer Assessment 1"
author: Johannes Marais
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The following code will set the working directory and load the data into a dataframe called activityData. The dataframe is then converted to a dplyr dataframe as that would make grouping the data easier for the next questions. When loading the data, only the collumn classes are changed as the Date is loaded as a factor by default and not a date. The NA values are kept as placeholder as they will be supstitituted in a later section.

```{r echo=TRUE,message=FALSE}
library(dplyr) #Load the dplyr library
setwd("~/Documents/Coursera/Specialization/05 Reproducable Research/Assignment/Assignment 1/RepData_PeerAssessment1") #Set the working directory to the location of the file
activityData<-read.csv("activity.csv",colClasses = c("integer","Date","integer")) #Load the data
activityData<-tbl_df(activityData) #Convert the data to a dplyr tbl class
str(activityData) #Display the data
```

## What is mean total number of steps taken per day?

```{r echo=TRUE}
library(ggplot2) #Load the ggplot2 library
dailySum<- activityData %>% group_by(date) %>%summarise(dailysum=sum(steps,na.rm=TRUE))
hist(dailySum$dailysum)
qplot(dailySum$dailysum,geom='histogram')
```

## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
